Exploration of Red Wine Quality by Jonathan Allen
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)
library(scales)
library(memisc)
library(gridExtra)
library(psych)
library(RColorBrewer)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('wineQualityReds.csv', row.names = 1)
```

This report contains an exploration of a dataset that contains ratings of 
1,599 red wines. Note that 1 dm^3 = 1 L. The variables are as follows 
(taken from Cortez et al., 2009):

####Input variables (based on physicochemical tests):
1. fixed acidity (tartaric acid - g / dm^3)
2. volatile acidity (acetic acid - g / dm^3)
3.citric acid (g / dm^3)
4. residual sugar (g / dm^3)
5. chlorides (sodium chloride - g / dm^3
6. free sulfur dioxide (mg / dm^3)
7. total sulfur dioxide (mg / dm^3)
8. density (g / cm^3)
9. pH
10. sulphates (potassium sulphate - g / dm3)
11. alcohol (% by volume)
####Output variable (based on sensory data): 
12. quality (score between 0 and 10)

####Description of attributes:
1. fixed acidity: most acids involved with wine or fixed or nonvolatile (do not
evaporate readily)
2. volatile acidity: the amount of acetic acid in wine, which at too high of 
levels can lead to an unpleasant, vinegar taste
3. citric acid: found in small quantities, citric acid can add 'freshness' and 
flavor to wines
4. residual sugar: the amount of sugar remaining after fermentation stops, it's 
rare to find wines with less than 1 gram/liter and wines with greater than 45 
grams/liter are considered sweet
5. chlorides: the amount of salt in the wine
6. free sulfur dioxide: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine
7. total sulfur dioxide: amount of free and bound forms of S02; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
8. density: the density of water is close to that of water depending on the 
percent alcohol and sugar content
9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
to 14 (very basic); most wines are between 3-4 on the pH scale
10. sulphates: a wine additive which can contribute to sulfur dioxide gas (S02) 
levels, wich acts as an antimicrobial and antioxidant
11. alcohol: the percent alcohol content of the wine
Output variable (based on sensory data): 
12. quality (score between 0 and 10)

# Univariate Plots Section

Dimensions of data set:
```{r echo=FALSE, message=FALSE, warning=FALSE, Dimensions}
#get the dimensions of the data frame
dim(df)
```

Structure of the data set:
```{r echo=FALSE, message=FALSE, warning=FALSE, Structure}
#get a summary of the data's structure
str(df)
```

Summary of the data set:
```{r echo=FALSE, message=FALSE, warning=FALSE, Summary}
#print a summary of all of the variables
summary(df)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Add_Quality_Factor}
#create a new column with 'quality' as a factor, instead of an int
df$quality.f <- as.factor(df$quality)
```

First of all, let's get a look at the distribution of quality scores for this
selection of wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
#bar plot to show distribution of quality scores
ggplot(data = df, aes(x = factor(quality))) +
  geom_bar() +
  xlab('quality')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#count/distribution of quality scores
table(df$quality)
```

This looks clean and makes sense. A rating scale like this has discrete values, 
and though the wines were rated on a 1-10 scale, the wines in the data set only 
contain values between 3 and 8. It also makes intuitive sense that there are a 
lot of decent wines, few excellent wines, and few bad wines.

Now let's look at histograms of the different chemical properties of the wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Fixed_Acidity}
#plot histogram of fixed acidity
ggplot(data = df, aes(x = fixed.acidity)) +
  geom_histogram() +
  xlab('fixed acidity (tartaric acid, g/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Fixed_Acidity2}
#summary of fixed acidity
summary(df$fixed.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Volatile_Acidity}
#plot histogram of volatile acidity
ggplot(data = df, aes(x = volatile.acidity)) +
  geom_histogram() +
  xlab('volatile acidity (acetic acid, g/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Volatile_Acidity2}
#summary of volatile acidity
summary(df$volatile.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Citric_Acid}
#plot histogram of citric acid
ggplot(data = df, aes(x = citric.acid)) +
  geom_histogram() +
  xlab('citric acid (g/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Citric_Acid2}
#summary of citric acid
summary(df$citric.acid)
```

The amount of tartaric acid (fixed acidity) in the wines is roughly a order of
magnitude greater (4.6-15.9 g/dm^3) than the amount of acetic acid (volatile
acidity) (0.12-1.58 g/dm^3) or citric acid (0.0-1.0 g/dm^3).

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_All_Acidity}
#want to plot all three measures of acidity together
#make a new data frame to pull out the necessary pieces

#intermediate data frame
df.i <- df[c('fixed.acidity', 'volatile.acidity', 'citric.acid')]

#gather with tidyr  
df.acids <- gather(data = df.i, "acidity", "g/dm^3", 
                   c('fixed.acidity', 'volatile.acidity', 'citric.acid'))

#plot hitogram for all fixed acidity, volatile acidity, and citric acid
ggplot(df.acids, aes(x = df.acids$`g/dm^3`)) +
  geom_histogram(aes(fill = acidity), alpha = 0.7, 
                 position = 'identity', binwidth = 0.1) +
  xlab('g/dm^3') +
  ggtitle('Three Types of Acidity')
```

Here we can see how the amounts of each type of acidity relate in frequency.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Residual_Sugar}
#histogram of residual sugar
ggplot(data = df, aes(x = residual.sugar)) +
  geom_histogram() +
  xlab('residual sugar (g/dm^3)')
```

```{r, echo=FALSE, Univariate_Residual_Sugar_Log}
#histogram of residual sugar, log_10 scale
ggplot(data = df, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 0.02) +
  scale_x_log10(breaks = c(1,2,4,8,16)) +
  geom_vline(aes(xintercept = 2.2, color = 'red'), show.legend = FALSE) +
  xlab('residual sugar (g/dm^3)')
```

```{r, echo=FALSE, Univariate_Residual_Sugar_Log2}
#summary of residual sugar
summary(df$residual.sugar)
```

Transformed the long tail data to get a better look at what is going on. The 
median is 2.2 g/dm^3, and has been marked in red. And even using a log_10 scale,
the data is still skewed to the right.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Chlorides}
#histogram of chlorides
ggplot(data = df, aes(x = chlorides)) +
  geom_histogram() +
  xlab('chlorides (sodium chloride, g/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Chlorides_Log}
#histogram of chlorides, log_10 scale
ggplot(data = df, aes(x = chlorides)) +
  geom_histogram(binwidth = 0.02) +
  scale_x_log10(breaks = c(.01, 0.02, 0.04, 0.08, 0.16, 0.32)) +
  geom_vline(aes(xintercept = 0.07900, color = 'red'), show.legend = FALSE) +
  xlab('chlorides (sodium chloride, g/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Chlorides_Log2}
#summary of chlorides
summary(df$chlorides)
```

Here again, I transformed the long tail data, which looks much closer to normal.
However, it still is skewed to the right. The median is 0.07900, which 
is marked in red.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Free_Sulfur_Dioxide}
#histogram of free sulfur dioxide
ggplot(data = df, aes(x = free.sulfur.dioxide)) +
  geom_histogram() +
  xlab('free sulfur dioxide (mg/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Free_Sulfur_Dioxide2}
#histogram of free sulfur dioxide on a log_10 scale
ggplot(data = df, aes(x = free.sulfur.dioxide)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_log10(breaks = c(1,2,4,8,16,32,64)) +
  xlab('free sulfur dioxide (mg/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Free_Sulfur_Dioxide3}
#summary of free sulfur dioxide
summary(df$free.sulfur.dioxide)
```

I transformed the data due to the long tail, and adjusted the binwidth. Though
intuitively, free sulfur dioxide would seem to be a continuous variable, the 
recorded values are all whole numbers (mg/dm^3), hence the distinctive bars.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Total_Sulfur_Dioxide}
#histogram of total sulfur dioxide
ggplot(data = df, aes(x = total.sulfur.dioxide)) +
  geom_histogram() +
  xlab('total sulfur dioxide (mg/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univ_Total_Sulfur_Dioxide2}
#histogram of total sulfur dioxide on a log_10 scale
ggplot(data = df, aes(x = total.sulfur.dioxide)) +
  geom_histogram() +
  scale_x_log10(breaks = c(8,16,32,64,128,256)) +
  xlab('total sulfur dioxide (mg/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univ_Total_Sulfur_Dioxide3}
#summary of total sulfur dioxide
summary(df$total.sulfur.dioxide)
```

Another long-tailed histogram, so I again transformed it to a log_10 scale. This
looks much more like a normal distribution. Most wines have a total sulfur
dioxide measurement of between 10 and 100 mg/dm^3.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Density}
#histogram of density
ggplot(data = df, aes(x = density)) +
  geom_histogram() +
  geom_vline(aes(xintercept = 1.0), color = 'blue', show.legend = FALSE) +
  xlab('density (g/cm^3)') +
  scale_x_continuous(breaks = c(0.99,0.9925,0.995,0.9975,1.0))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Density2}
#summary of density
summary(df$density)
```

I've added a blue vertical line to mark the density of water. The majority of
the wines are very slighlty less dense. For example, the mean density of the 
wines is 0.9967 g/cm^3, compared to the 1.0 g/cm^3 density of water.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_pH}
#histogram of pH
ggplot(data = df, aes(x = pH)) +
  geom_histogram() +
  xlab('pH')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_pH2}
#summary of pH
summary(df$pH)
```

Almost all of the wines have a pH between 3 and 4, which are the acidities of 
orange juice and tomato juice, respectively (2). Keep in mind that an acidity of
7 considered neutral (neither acid nor base), and is the acidity of pure water.
Another point of note is that the pH scale is a log_10 scale, meaning that a pH 
3.0 is ten times more acidic than a pH of 4.0.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Sulphates}
#histogram of sulphates
ggplot(data = df, aes(x = sulphates)) +
  geom_histogram() +
  xlab('sulphates (potassium suphate, g/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Sulphates2}
#histogram of sulphates on a log scale
ggplot(data = df, aes(x = sulphates)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_log10(breaks = c(0.4, 0.566, 0.8, 1.131, 1.6)) +
  xlab('sulphates (potassium suphate, g/dm^3)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Sulphates3}
#summary of sulphates
summary(df$sulphates)
```

Transforming the long-tailed graph does look a little better. Almost all of the
wines have fewer than 1 gram of suphates per dm^3, but there is at least one
wine with twice that.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Alcohol}
#histogram of % alcohol
ggplot(data = df, aes(x = alcohol)) +
  geom_histogram() +
  xlab('alcohol (% by volume)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Alcohol2}
#summary of % alcohol
summary(df$alcohol)
```

Most of the wines have an alcohol content between 9 and 12 percent. The plot
is also skewed to the right.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Buckets}
#create five buckets for % alcohol, to turn it into a categorical variable
buckets = quantile(df$alcohol, c(0, .20, .40, .60, .80, 1.00))
df$alcohol.bucket <- cut(df$alcohol, breaks = buckets)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Balance}
#create new variable called 'balance' = fixed acidity / residual sugar
df$balance <- df$fixed.acidity / df$residual.sugar

#histogram of new variable, balance
ggplot(data = df, aes(x = balance)) +
  geom_histogram() +
  xlab('balance (fixed acidity / residual sugar)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Balance_Summary}
#summary of balance
summary(df$balance)
```

Most of the wines have 2.5-5.0 times as much tartaric acid (fixed acidity) as 
residual sugar.

# Univariate Analysis

### What is the structure of your dataset?

There are 1,599 wines in the data set. Each wine has been rated on a 1-10 scale 
(the recorded score is the median of at least 3 ratings given by wine experts) 
and has been measured in 11 different chemical attributes. The only true
categorical variable is the wine's quality, with 10 being the best wine and 1
being the worst.

More than 80% of the wines have a rating of either 5 or 6.

### What is/are the main feature(s) of interest in your dataset?

From research about wine, it is clear that the amount of acidity itself, or the
amount of residual sugar itself is not an indicator of a high quality wine. What
matters is 'balance' (3). So, one major feature of interest is to see whether a
certain ratio of acidity (my guess is that fixed acidity is what to look at) to
residual sugar correlates with a higher quality wine.

So, I anticipate that balance (fixed acidity / residual sugar) and quality are 
the main features of interest.

### What other features in the dataset do you think will help support your \ investigation into your feature(s) of interest?

I think that alcohol content, volatile acidity, and citric acid will also play a
part in refining the model. I'll also check the other variables, but I
anticipate them having less of a role.

Since density varies so little betweeen samples, and it has a moderate
correlation with both fixed acidity and alcohol, I will not be looking at
density more closely.

Similarly, pH has a moderate negative correlation with fixed acitity and citric
acid, and a positive correlation with volatile acidity. And overall, pH measures
the same thing, acidity. With so many measures of acidity, I think it's safe to 
pull pH out and keep the other three measures of acidity, since they are more 
specific.

### Did you create any new variables from existing variables in the dataset?

I created a variable called 'balance', which is the ratio between fixed acidity
and residual sugar.

I also broke up the wines into buckets based on alcohol percentage, split at 
20th percentile marks.

Lastly, I created a copy of the 'quality' column, with a datatype of factor.
This was done largely for convenience, so I could do any desired mathematics off
the one, and use the other as a categorical variable more easily.

### Of the features you investigated, were there any unusual distributions? \ Did you perform any operations on the data to tidy, adjust, or change the form \ of the data? If so, why did you do this?

There were a number of initial histograms that were skewed to the right (none 
were skewed to the left), and so I used log transformations on them. These were
residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, and
sulphates. I did so to get a better look at what was happening at the peaks.

I also noticed that, though the distributions of the chemical properties of the
wines would intuitively be continuous, as the binwidth got smaller, they
started to look discrete, due to the number of samples, and the rounding of the
recorded data.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Scatter_Plot_Matrix}
#df2 leaves out density and pH
#c(1:7, 10:12, 15)
df2 <- df[c("fixed.acidity", "volatile.acidity", 
                              "citric.acid", "residual.sugar", 
                              "chlorides", "free.sulfur.dioxide",
                              "total.sulfur.dioxide", "sulphates",
                              "alcohol", "quality", "balance")]
#scatter plot matrix of variables in df2
pairs.panels(df2)
names(df)
```

Wow! My calculation of 'balance' has almost no correlation to quality. There
goes my bright idea.

Also surprising to me, the strongest correlation indicated from this scatterplot
matrix is alcohol % (moderate positive correlation), followed by volatile 
acidity (weak negative correlation). The latter was not surprising.

From here, I want to look at the relationship between quality and some of the
other variables, particularly those with a higher correlation with quality. I
would also like to look at some of the correlations between variables that are
*not* quality, to get a better picture of what's going on between them.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_Balance}
#box plot of balance vs quality
ggplot(data = df, aes(x = quality.f, y = balance)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('Balance')
```

Yep. Looks like no correlation.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Box_Plots}
#box plot of % alcohol vs quality
ggplot(data = df, aes(x = quality.f, y = alcohol)) +
  geom_boxplot() +
  ylab('% Alcohol') +
  xlab('Quality Rating')
```

Here, the moderate positive correlation between % alcohol and quality is easy
to see. One thing that stands out is the smaller interquartile range of the
quality rating of 5. I wonder if 9.5-10% alcohol is fairly easy to achieve for
wine, such that many wines fall there. Along with rating 6, this rating includes
80% of all the wines in the data set, so that's a very highly-populated and 
small interquartile range.

```{r echo=FALSE, message=FALSE, warning=FALSE, Biv_Sulphates}
#box plot of sulphates vs quality
ggplot(data = df, aes(x = quality.f, y = sulphates)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('Sulphates')
```

There are a lot of outliers on this plot, but you can still see the weak 
positive correlation.

```{r echo=FALSE, message=FALSE, warning=FALSE, Biv_Sulphates_Zoom}
#zoomed in box plot of sulphates vs quality
ggplot(data = df, aes(x = quality.f, y = sulphates)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('Sulphates') +
  coord_cartesian(ylim = c(0.35,1.1))
```

There, it's much easier to see after zooming in, particularly if you look at the
median lines. But, all those outliers make for a weaker correlation.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_vs_Residual_Sugar}
#scatter plot of residual sugar vs % alcohol
ggplot(data = df, aes(x = alcohol, y = residual.sugar)) +
  geom_point(alpha = 1/3) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(9,14,1)) +
  xlab('% Alcohol') +
  ylab('Residual Sugar')
```

Residual sugar and alcohol actual have almost zero correlation (0.04 correlation
coefficient, according to the scatterplot matrix), which this plot corroborates.
Though I was hoping to find a stronger correlation, this result does makes sense
in light of the fact that different grapes have differing sugar content, which
means that different wines start with different levels of sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivarite_Acids}
#scatter plot of volatile acidity vs fixed acidity
a1 = ggplot(data = df, aes(x = fixed.acidity, y = volatile.acidity)) +
  geom_point(alpha = 1/5) +
  xlab('Fixed Acidity') +
  ylab('Volatile Acidity') +
  scale_x_continuous(breaks = seq(6,16,2))

#scatter plot of citric acid vs fixed acidity
a2 = ggplot(data = df, aes(x = fixed.acidity, y = citric.acid)) +
  geom_point(alpha = 1/5) +
  xlab('Fixed Acidity') +
  ylab('Citric Acid') +
  scale_x_continuous(breaks = seq(6,16,2))

#scatter plot of citric acid vs volatile acidity
a3 = ggplot(data = df, aes(x = volatile.acidity, y = citric.acid)) +
  geom_point(alpha = 1/5) +
  xlab('Volatile Acidity') +
  ylab('Citric Acid')

#plot all three
grid.arrange(a1, a2, a3, ncol = 1)
```

It's interesting that the acidities correlate in different directions. I would
have thought that more acidity would have meant higher values in all categories
of acidity. However, fixed acidity and volatile acidity have a weak negative 
correlation, which wasn't intuitive to me. Fixed acidity and citric acid have a 
strong positive correlation. And citric acid and volatile acidity have a 
moderate negative correlation.

This second relationship makes sense, thinking back to the description of the 
attributes. Citric acid gives 'freshness' and flavor, which is generally good, 
whereas too much volatile acidity results in an 'unpleasant, vinegar taste'.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multi_Plots_Volatile_Acidity}
#box plot of volatile acidity vs quality
ggplot(data = df, aes(x = quality.f, y = volatile.acidity)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('Volatile Acidity')
```

Here, you can see the weak (almost moderate) negative correlation between
volatile acidity and quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_Citric_Acid}
#box plot of citric acid vs quality
ggplot(data = df, aes(x = quality.f, y = citric.acid)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('Citric Acid')
```

And this is the weak positive relationship between quality and citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Total_Sulfur_Dioxide}
#box plot of total sulfur dioxide vs quality
ggplot(data = df, aes(x = quality.f, y = total.sulfur.dioxide)) +
  geom_boxplot() +
  xlab('quality')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Biv_Total_Sulfur_Dioxide_Zoom}
#zoomed in box plot of total sulfur dioxide vs quality
ggplot(data = df, aes(x = quality.f, y = total.sulfur.dioxide)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0,160)) +
  xlab('Quality') +
  ylab('Total Sulfur Dioxide')
```

Box plots showing the weak negative correlation between total sulfur dioxide
and quality. Zoomed in on the second plot. Interestingly, the lowest quality 
wine also has the lowest total sulfur dioxide. That might be why the correlation 
isn't stronger.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivarite_Sulfur_Dioxide}
#box plot of free sulfur dioxide vs quality
ggplot(data = df, aes(x = quality.f, y = free.sulfur.dioxide)) +
  geom_boxplot() +
  geom_hline(yintercept = 50, color = 'red') +
  xlab('Quality') +
  ylab('Free Sulfur Dioxide')
```

The description of the attributes mentions that concentrations of over 50 ppm 
free SO2 (free sulfur dioxide), its presence becomes evident in the "nose and 
taste" (this level is marked by a red horizontal line). From this plot, it's 
hard to tell whether that's a good or bad thing. There is supposedly a weak 
negative correlation between free sulfur dioxide and quality, but the plot makes
me less confident in that, since on both the high and low quality ends, there 
tends to be lower levels of free sulfur dioxide.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \ investigation. How did the feature(s) of interest vary with other features in \ the dataset?

Alcohol was moderately correlated with quality (0.48), which I found surprising. 
The second strongest correlation was a weak negative correlation between 
volatile acidity and quality (-.39). That second relationship was not 
surprising, since wine that tastes like vinegar sounds awful.

Another surprise was that my created variable of 'balance' had no correlation
with quality. I know that the concept of balance is very important in wine, so
I think there must be more to it than my variable took into account.

### Did you observe any interesting relationships between the other features \ (not the main feature(s) of interest)?

The relationship between the acidities was interesting. Though one might expect
them to all correlate together, that was not the case. Rather, fixed acidity
(tartaric acid) and citric acid had a strong positive correlation, and volatile
acidity (acetic acid) had a negative correlation with both (weak vs fixed
acidity and moderate vs citric acid).

### What was the strongest relationship you found?

The strongest relationships that I found with quality were alcohol (0.48 
correlation coefficient) and then volatile acidity (-0.39 correlation 
coefficient). After that, sulphates had a weak positive correlation with quality
(0.25), as did citric acid (0.23).

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Melt_for_Balance_Acids}
#want long form of just quality.f, and citric acid + volatile acidity
#get just the necessary columns
df.i <- df[c('quality.f', 'citric.acid', 'volatile.acidity')]
#gather into long form
df.acids <- gather(df.i, acid.type, acidity, c('citric.acid','volatile.acidity'))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Acids_and_Quality}
#box plot of citric acid vs quality and volatile acidity vs quality
ggplot(data = df.acids, aes(x = quality.f, y = acidity, fill = acid.type)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('g/dm^3') +
  guides(fill=guide_legend(title=''))
```

Now *that's* interesting. As the wines get better, the amount of citric acid and
acetic acid (i.e. volatile acidity) tend to be nearly equal, as the amount of 
acetic acid decreases, and the amount of citric acid increases. Is this what 
'balance' *should* mean?

```{r echo=FALSE, message=FALSE, warning=FALSE, Balanced_Acidity}
#create a new variable, 'balance.acidity' = citric acid / volatile acidity
df$balance.acidity <- df$citric.acid / df$volatile.acidity
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Balanced_Acidity_Plot}
#box plot of balance.acidity vs quality
ggplot(data = df, aes(x = quality.f, y = balance.acidity)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('Citric Acid / Volatile Acidity')
```

This looks somewhat promising. What's the correlation coefficient?

```{r echo=FALSE, message=FALSE, warning=FALSE, Balance_Correlation}
#calculate correlation coefficient of balance.acidity to quality
cor(df$quality, df$balance.acidity)
```

Hmm. With a correlation coefficient of 0.32, that's still a weak [positive] 
correlation. It's slightly better than citric acid and volatile acidity are on 
their own, but it's no home run.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multi_Alcohol_Balance_Quality}
#scatter plot of balance.acidity vs quality, with alcohol.bucket as color
ggplot(data = subset(df, !is.na(df$alcohol.bucket)), 
       aes(x = quality, y = balance.acidity)) +
  geom_point(aes(color = alcohol.bucket), position = 'jitter') +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_hline(yintercept = 1, color = 'black') +
  scale_color_brewer(type = 'seq', palette = 3) +
  xlab('Quality') +
  ylab('Citric Acid / Acetic Acid') +
  guides(color=guide_legend(title='% Alcohol'))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Multi_Alcohol_Balance_Quality_2}
#scatter plot of balance.acidity vs quality, with alcohol.bucket as color
#zoomed in
ggplot(data = subset(df, !is.na(df$alcohol.bucket)), 
       aes(x = quality, y = balance.acidity)) +
  geom_point(aes(color = alcohol.bucket), position = 'jitter') +
#  scale_y_log10() +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_hline(yintercept = 1, color = 'black') +
  scale_color_brewer(type = 'seq', palette = 3) +
  xlab('Quality') +
  ylab('Citric Acid / Acetic Acid') +
  guides(color=guide_legend(title='% Alcohol')) +
  coord_cartesian(ylim = c(0,2))
```

Here we see balance (citric acid/acetic acid) versus quality, with % alcohol 
buckets added via color. The second plot is zoomed in to get a better look at 
what is happening near y = 1.0. The low number of data points at the extremes
makes it harder to see trends. However, between quality ratings 5 and 7, it's
clear that there's a trend that the quality of the wine improves as the balance
moves from a lower to higher ratio of citric acid to acetic acid. Most of the
points with quality 5 are well below the 1.0 line, quality 6 points are more
evenly spread above and below, and wines with quality 7 are predominantly above
the 1.0 line.

```{r echo=FALSE, message=FALSE, warning=FALSE, Building_Linear_Model}
#building a linear model
#adding pieces in order of descending strength of correlation
m1 <- lm(I(quality) ~ I(alcohol), data = df) #start with strongest correlation
m2 <- update(m1, ~ . + balance.acidity) #add in balance.acidity, next
m3 <- update(m2, ~ . + sulphates) #add in sulphates (next strongest correlation)
m4 <- update(m3, ~ . + total.sulfur.dioxide) #add total sulfur dioxide
m5 <- update(m4, ~ . + chlorides) #add in chlorides
#these (below) didn't make the model any better
# m6 <- update(m5, ~ . + fixed.acidity) #add in fixed acidity
# m7 <- update(m6, ~ . + residual.sugar)
# m8 <- update(m7, ~ . + free.sulfur.dioxide)
mtable(m1, m2, m3, m4, m5)
```

The R-sqaured value in models m1-m5 vary from 0.227 to 0.326, so I wouldn't 
trust the model very far at all. I originally tried an m6, m7 and m8, adding in 
fixed acidity, residual sugar, and free sulfur dioxide, but they didn't improve
the model at all (the R-squared value didn't change), so I'm leaving them out.
If two models are equally accurate, choose the simpler one.

I wonder whether combining volatile acidity and citric acid made for a stronger
or weaker model?

```{r echo=FALSE, message=FALSE, warning=FALSE, Building_Linear_Model2}
#building another linear model
#adding pieces in order of descending strength of correlation
m1 <- lm(I(quality) ~ I(alcohol), data = df) #start with strongest correlation
m2 <- update(m1, ~ . + volatile.acidity + citric.acid) #use separate acids
m3 <- update(m2, ~ . + sulphates) #add in sulphates (next strongest correlation)
m4 <- update(m3, ~ . + total.sulfur.dioxide) #add total sulfur dioxide
m5 <- update(m4, ~ . + chlorides) #add in chlorides
#these (below) didn't make the model any better
# m6 <- update(m5, ~ . + fixed.acidity) #add in fixed acidity
# m7 <- update(m6, ~ . + residual.sugar)
# m8 <- update(m7, ~ . + free.sulfur.dioxide)
mtable(m1, m2, m3, m4, m5)
```

Combining volatile acidity and citric acid into a 'balance' number actually
weakened the model. Here, when using them separately, we can get up to an
R-squared of 0.352, an improvement of 0.026. Still not something I would rely
upon, though.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \ investigation. Were there features that strengthened each other in terms of \ looking at your feature(s) of interest?

The most interesting relationship that I observed was when plotting both citric
acid vs quality and volatile acidity vs quality at the same time. It became
clear that poor wines have much more volatile acidity than citric acid, and
better wines, on average, tend to have more equal amounts.

I thought that citric acid and volatile acidity might strengthen each other
when put together as a calculation of 'balance', but they actually weakend
their relationships with quality.

Another interesting relationship was between alcohol and quality. I didn't
expect to see a positive correlation between them. But, it was the strongest
correlation of the bunch.

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \ strengths and limitations of your model.

I did create a linear model with the dataset, but it wasn't strong. The
R-value topped out at 0.352. From these chemical measurements, the quality of
wine does not seem to be very predictable.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
#bar plot showing distribution of quality scores
ggplot(data = df, aes(x = factor(quality))) +
  geom_bar() +
  ggtitle('Distribution of Quality Ratings') +
  ylab('Count') +
  xlab('Quality Rating')
```

### Description One
This distribution shows that the vast majority of wines were given a quality 
rating of 5 or 6. In particular, this highlights one of the limitations of the
data set, which is that there were so few wines at the extremes, that patterns
were difficult to detect, there.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
#box plot of citric acid vs quality and volatile acidity vs quality
ggplot(data = df.acids, aes(x = quality.f, y = acidity, fill = acid.type)) +
  geom_boxplot() +
  xlab('Quality') +
  ylab('g/dm^3') +
  guides(fill=guide_legend(title='')) +
  ggtitle('Interaction of Volatile Acidity and Citric Acid over Quality')
```

### Description Two
Wines with lower ratings tend to have much higher levels of volatile acidity
compared to the levels of citric acid. As wine quality improves, these two
levels tend to approach one another, and for qualities of 7 and 8, particularly,
it's clear that there is roughly a balance between the two acidities.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
#zoomed in scatter plot of balance.acidity vs quality, with 
#alcohol.bucket as color
ggplot(data = subset(df, !is.na(df$alcohol.bucket)), 
       aes(x = quality, y = balance.acidity)) +
  geom_point(aes(color = alcohol.bucket), position = 'jitter') +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_hline(yintercept = 1, color = 'black') +
  scale_color_brewer(type = 'seq', palette = 3) +
  xlab('Quality') +
  ylab('Citric Acid / Acetic Acid') +
  ggtitle('Citric Acid/Acetic Acid by Quality and % Alcohol') +
  guides(color=guide_legend(title='% Alcohol')) +
  coord_cartesian(ylim = c(0,2))
```

### Description Three
Higher quality ratings tend to be associated with higher percentages of alcohol,
as well as higher levels of citric acid relative to volatile acidity. However,
the data does not seem to support a predictive model.

# Reflection

The 'wineQualityReds' data set has information on 1,599 different wines, including
ratings that are an average of at least three ratings given by wine experts. To
begin, I took a look at variables individually, and then began to look for
interesting interactions, and to find which variables had a greater impact on
quality.

Wine is complex. The chemical attribute measurements of the wines clearly did
not encompass all of the things that the wine experts used to make their rating.
Even with a data set of 1,599 wines, the strength of detected patterns were 
moderate, at best. Of the attributes in the data set, percent alcohol was the 
strongest indicator of wine quality of the attributes measured. Another weak 
indicator was how much volatile acidity the wine contained (negative 
correlation), which had  an intriguing relationship with the amount of citric 
acid. Better wines tended to have higher levels of citric acid, relative to the 
amount of volatile acidity. I tried to fit a linear model to the data, but it
only accounted for 35.2% of the variance in the quality rating in this data set.

One limitation of the data set is that there were so few wines at the more
extreme ratings (1-3, 8-10). This compresses the data into a small number of
quality values, making it more difficult to distinguish differences and 
patterns. Another limitation is that quality ratings are completely subjective.
Yes, they were an average of at least three ratings, but those individual
ratings are based on the thoughts of an individual person, which cannot be
objectively measured. For further investigation, I would try to obtain
measurements of tannins, a component of wine that, when researching, I found
to be quite important (4). I would also try to get a better understanding of
what things make up the concept of 'balance' in a wine, which may include
attributes not measured in this data set.

Overall, I'm not convinced that statistical tests are that helpful regarding
finding high quality wines, since personal preference likely plays a large
role in rating. However, measurements and statistics *would* be useful in
classifying wines (sweet, dry, etc.). For example, white wine and red wine might 
have different chemical tendencies which would be another avenue of further 
inquiry. Also, statistics could possibly help in finding wines where chemical
attributes are far outisde of the recommended range, and thereby assist in
screening bad wines. But, I don't truly anticipate boiling the quality of wine
down to a statistical model.